### 2021-04-08 papers/SYRCoSE/future.tex
Typos and British spelling
find . -name "*.tex" -maxdepth 1 -exec aspell --lang=en_GB --mode=tex check "{}" \;

First, we plan to apply our library to a wider range of problems. 
We are going to develop a mechanized semantics of some classical 
languages used to model concurrency, in particular 
the calculus of communicating systems~(CCS)~\cite{Milner:80} and 
$\pi$-calculus~\cite{Milner:99}.      
We also plan to continue our work on 
expressing various relaxed models 
of shared memory~\cite{Lahav-al:POPL16, Lahav-al:PLDI17, Podkopaev-al:POPL19}
in terms of event structures.

Second, we want to cover other classes of event structures in our library, 
in particular bundle~\cite{Langerak:91}, flow~\cite{Boudol-Castellani:1991}
and stable~\cite{Winskel:82, Winskel:86} event structures.
We plan to use them to develop mechanized \emph{denotational} semantics 
of concurrent languages and relaxed shared memory models~\cite{Dodds-al:ESOP18}.

Finally, we plan to mechanize in Coq classical results 
that connect various classes of event structures%  
~\cite{Nielsen-al:1981, Boudol-Castellani:1991}. 
It would allow us to easily establish the connection between operational and denotational 
semantics of concurrent languages.

>>>>

First, we plan to apply our library to a wider range of problems. 
We are going to develop a mechanised semantics of some classical 
languages used to model concurrency, in particular 
the calculus of communicating systems~(CCS)~\cite{Milner:80} and 
$\pi$-calculus~\cite{Milner:99}.      
We also plan to continue our work on 
expressing various relaxed models 
of shared memory~\cite{Lahav-al:POPL16, Lahav-al:PLDI17, Podkopaev-al:POPL19}
in terms of event structures.

Second, we want to cover other classes of event structures in our library, 
in particular bundle~\cite{Langerak:91}, flow~\cite{Boudol-Castellani:1991}
and stable~\cite{Winskel:82, Winskel:86} event structures.
We plan to use them to develop mechanised \emph{denotational} semantics 
of concurrent languages and relaxed shared memory models~\cite{Dodds-al:ESOP18}.

Finally, we plan to mechanise in Coq classical results 
that connect various classes of event structures%  
~\cite{Nielsen-al:1981, Boudol-Castellani:1991}. 
It would allow us to easily establish the connection between operational and denotational 
semantics of concurrent languages.

### 2021-04-08 papers/SYRCoSE/intro.tex
Typos and British spelling
find . -name "*.tex" -maxdepth 1 -exec aspell --lang=en_GB --mode=tex check "{}" \;

The main advantage of the event structures 
compared to traditional interleaving semantics 
is that they give a more compact and consice 
representation of progams' behavior.
For example, consider the following code snippet.

@@

The same information can be encoded in a single 
event structure containing $6$ events in total
(see \cref{fig:intro-es}). 
In the event structure there are two types of edges 
between the events. The gray arrows $e_1 \arrowCA e_2$ 
represent the \emph{causality relation}, a 
partial order reflecting the causal relationship
between the atomic events of computation.
The red edges $e_1 \arrowCF e_2$ represent 
the \emph{conflict relation} which is 
a symmetric and irreflexive relation 
encoding mutually exclusive events.
Each particular trace can be extracted from the event structure
as a linearisation of some \emph{configuration}, 
that is a causally-closed and conflict-free 
subset of events.

@@

In the programming language and formal semantics research communities 
it becomes the de-facto standard to complement the theoretical 
studies with their mechanization in the \emph{proof assistants}
like Coq, Agda, Isabelle, Arend, and others,
as this process increases the reliability and reproducibility 
of the scientific results.
Yet, as far as we know, there is a little work on 
mechanization of event structures theory.

Our work is aimed to close this gap. 
Our goal is to develop a Coq library containing 
a comprehensive set of common definitions, lemmas, 
and tactics that would allow the researchers 
to utilize the theory of event structures 
for the needs of their domain.

In this work in progress report we sketch 
the common design princples behind our library
and give a concrete example of its usage  
by developing a formal mechanized semantics of simple 
register machine with shared memory.

>>>>

The main advantage of the event structures 
compared to traditional interleaving semantics 
is that they give a more compact and concise 
representation of programs' behaviour.
For example, consider the following code snippet.

@@

The same information can be encoded in a single 
event structure containing $6$ events in total
(see \cref{fig:intro-es}). 
In the event structure there are two types of edges 
between the events. The grey arrows $e_1 \arrowCA e_2$ 
represent the \emph{causality relation}, a 
partial order reflecting the causal relationship
between the atomic events of computation.
The red edges $e_1 \arrowCF e_2$ represent 
the \emph{conflict relation} which is 
a symmetric and irreflexive relation 
encoding mutually exclusive events.
Each particular trace can be extracted from the event structure
as a linearisation of some \emph{configuration}, 
that is a causally-closed and conflict-free 
subset of events.

@@

In the programming language and formal semantics research communities 
it becomes the de-facto standard to complement the theoretical 
studies with their mechanisation in the \emph{proof assistants}
like Coq, Agda, Isabelle, Arend, and others,
as this process increases the reliability and reproducibility 
of the scientific results.
Yet, as far as we know, there is a little work on 
mechanisation of event structures theory.

Our work is aimed to close this gap. 
Our goal is to develop a Coq library containing 
a comprehensive set of common definitions, lemmas, 
and tactics that would allow the researchers 
to utilise the theory of event structures 
for the needs of their domain.

In this work in progress report we sketch 
the common design principles behind our library
and give a concrete example of its usage  
by developing a formal mechanised semantics of simple 
register machine with shared memory.

### 2021-04-08 papers/SYRCoSE/overview.tex
Typos and British spelling
find . -name "*.tex" -maxdepth 1 -exec aspell --lang=en_GB --mode=tex check "{}" \;

We build our mechanization on top of \mathcomp~\cite{Mahboubi-Tassi:MATHCOMP17} library 
using the \ssreflect~\cite{Gonthier-al:SSR2016} extension of the \coq system.
We also use the \emph{small-scale reflection} 
methodology~\cite{Gonthier-Assia:SSR2010, Gonthier-al:SSR2016}, 
a key ingredient of the \ssreflect.

The small-scale reflection approach is based on 
pervasive use of symbolic representation of the proof goal, 
which \emph{relfects} the logical representation. 
The symbolic representation can be manipulated 
by the computational engine of the language, 
allowing the user to automate low-level routine 
proof management by using various decision 
and simplification procedures. 
Whenever the user needs to guide the proof 
it can switch to logical representation
and perform some proof steps manually.

@@

We also favor the computational encoding of semantics. 
Similarly to the recent related works on mechanization 
of operational semantics~\cite{Xia-al:POPL2019, Letan-al:CPP2020, Affeldt-al:ICMPC2019}, 
we encode the semantics as monadic interpreters.  
This allows us to extract~\cite{Letouzey:CCE2008} 
the semantics as a functional program and run it. 
We believe that the possibility to run the semantics 
is a very useful feature, as it allows 
to debug the formal semantics
and helps to develop better intuition about it.

>>>>

We build our mechanisation on top of \mathcomp~\cite{Mahboubi-Tassi:MATHCOMP17} library 
using the \ssreflect~\cite{Gonthier-al:SSR2016} extension of the \coq system.
We also use the \emph{small-scale reflection} 
methodology~\cite{Gonthier-Assia:SSR2010, Gonthier-al:SSR2016}, 
a key ingredient of the \ssreflect.

The small-scale reflection approach is based on 
pervasive use of symbolic representation of the proof goal, 
which \emph{reflects} the logical representation. 
The symbolic representation can be manipulated 
by the computational engine of the language, 
allowing the user to automate low-level routine 
proof management by using various decision 
and simplification procedures. 
Whenever the user needs to guide the proof 
it can switch to logical representation
and perform some proof steps manually.

@@

We also favour the computational encoding of semantics. 
Similarly to the recent related works on mechanisation 
of operational semantics~\cite{Xia-al:POPL2019, Letan-al:CPP2020, Affeldt-al:ICMPC2019}, 
we encode the semantics as monadic interpreters.  
This allows us to extract~\cite{Letouzey:CCE2008} 
the semantics as a functional program and run it. 
We believe that the possibility to run the semantics 
is a very useful feature, as it allows 
to debug the formal semantics
and helps to develop better intuition about it.

### 2021-04-08 papers/SYRCoSE/regmachine-eventstruct.tex
Typos and British spelling
find . -name "*.tex" -maxdepth 1 -exec aspell --lang=en_GB --mode=tex check "{}" \;

In this section we present operational semantics
which constructs a prime event structure encoding a 
behavior of the register machine.

@@

The first component $\eventSeq$ is a sequence of events 
$e_1 \identOrdDual \dots \identOrdDual e_n$
in the order reverse to the the one 
in which events are added to the structure. 
The second component is labeling function $\lLAB : \eventSet \fun \Lab$, 
assigning a label to each event.

@@

We use the inverse covering function because it is 
more convinient in our setting. Indeed, the semanics adds new event
at each step. Then it is convinient to require that in addition 
the small-step relation is provided with the $\lPO$ and $\lRF$
predecessors of a new event.

@@

The primitive conflict relation is generated by the $\lfPO$ function.
The two events are considered to be in primitive conflict if they are
not equal and have a commond $\lPO$ predecessor. 
For this definition to work properly, we also need 
to assume that in each thread has a special initial event 
labeled by a distinguished \emph{thread start} label $\tslab{}$.

@@

The construction starts from an initial event structure,
containing for each thread an event labeled by $\tslab{}$.
Below the label we depict the corresponding thread state.
Initially, each event is mapped to an initial thread state, 
consisting of an instruction pointer pointing to the first 
instruction to execute, and an initial mapping of registers,
denoted as $\bot$.
The first step executes the store instruction from 
the leftmost thread, and exits the program, 
since the execution of this thread terminates
(we omit $\exitInst$ instructions at the end of each thread for brevity).
Next the store from the rightmost thread is executed and corresponding
write event is added to the structure. 
After that the load instruction from the middle thread is executed. 
Since there are two matching write events in the event structure, 
two conflicting reads are conjoined to the event structure. 
Note that the events can be added non-deterministically in any order
respecting the causality. We could have first execute the rightmost thread
and add write $\wlab{x}{2}$ before $\wlab{x}{1}$, 
or add the read with label $\rlab{x}{2}$ 
before another read $\rlab{x}{1}$.

@@

The rules of operational semantics constructing 
the event structure are presented on~\cref{fig:eventstruct-sem}.
The first auxiliary rule \ESAddEventRule adds a new event, sets its 
label, $\lPO$ and $\lRF$ predecessors. 
The \ESIdleRule handles the case when a thread of 
the register machine performs an internal step 
without any side effect. 
It chooses an event $e$ together with 
the thread state $s$ corresponding to it
and performs one step reduction to a new state $s'$.
It then updates the mapping of events to thread states.   
The last two rules \ESStoreRule and \ESLoadRule
correspond to store and load perfromed by some thread.  
Similarly to \ESIdleRule, an event $e_{\lPO}$ is selected
and one reduction is performed from the corresponding thread state $s$.
Unlike the \ESIdleRule case, however, a new event $e$ is also generated.
In case of \ESLoadRule additionally an event $e_{\lRF}$ is selected,
such that it has a write label matching the read label of the new event.

>>>>

In this section we present operational semantics
which constructs a prime event structure encoding a 
behaviour of the register machine.

@@

The first component $\eventSeq$ is a sequence of events 
$e_1 \identOrdDual \dots \identOrdDual e_n$
in the order reverse to the the one 
in which events are added to the structure. 
The second component is labelling function $\lLAB : \eventSet \fun \Lab$, 
assigning a label to each event.

@@

We use the inverse covering function because it is 
more convenient in our setting. Indeed, the semantics adds new event
at each step. Then it is convenient to require that in addition 
the small-step relation is provided with the $\lPO$ and $\lRF$
predecessors of a new event.

@@

The primitive conflict relation is generated by the $\lfPO$ function.
The two events are considered to be in primitive conflict if they are
not equal and have a common $\lPO$ predecessor. 
For this definition to work properly, we also need 
to assume that in each thread has a special initial event 
labelled by a distinguished \emph{thread start} label $\tslab{}$.

@@

The construction starts from an initial event structure,
containing for each thread an event labelled by $\tslab{}$.
Below the label we depict the corresponding thread state.
Initially, each event is mapped to an initial thread state, 
consisting of an instruction pointer pointing to the first 
instruction to execute, and an initial mapping of registers,
denoted as $\bot$.
The first step executes the store instruction from 
the leftmost thread, and exits the program, 
since the execution of this thread terminates
(we omit $\exitInst$ instructions at the end of each thread for brevity).
Next the store from the rightmost thread is executed and corresponding
write event is added to the structure. 
After that the load instruction from the middle thread is executed. 
Since there are two matching write events in the event structure, 
two conflicting reads are conjoined to the event structure. 
Note that the events can be added non-deterministically in any order
respecting the causality. We could have first execute the rightmost thread
and add write $\wlab{x}{2}$ before $\wlab{x}{1}$, 
or add the read with label $\rlab{x}{2}$ 
before another read $\rlab{x}{1}$.

@@

The rules of operational semantics constructing 
the event structure are presented on~\cref{fig:eventstruct-sem}.
The first auxiliary rule \ESAddEventRule adds a new event, sets its 
label, $\lPO$ and $\lRF$ predecessors. 
The \ESIdleRule handles the case when a thread of 
the register machine performs an internal step 
without any side effect. 
It chooses an event $e$ together with 
the thread state $s$ corresponding to it
and performs one step reduction to a new state $s'$.
It then updates the mapping of events to thread states.   
The last two rules \ESStoreRule and \ESLoadRule
correspond to store and load performed by some thread.  
Similarly to \ESIdleRule, an event $e_{\lPO}$ is selected
and one reduction is performed from the corresponding thread state $s$.
Unlike the \ESIdleRule case, however, a new event $e$ is also generated.
In case of \ESLoadRule additionally an event $e_{\lRF}$ is selected,
such that it has a write label matching the read label of the new event.

### 2021-04-08 papers/SYRCoSE/regmachine.tex
Typos and British spelling
find . -name "*.tex" -maxdepth 1 -exec aspell --lang=en_GB --mode=tex check "{}" \;

For our case study we use a simple idealized model 
of a register machine, which consists of a finite 
sequence of instructions, an instruction pointer 
and an infinite set of registers. 
The syntax of the machine's language is 
shown on~\cref{fig:regmachine-syntax}.

@@

The semantics is given in a form of 
\emph{labeled transition system}~%
\footnote{As we have mentioned, in our \coq development we actually use 
the monadic encoding of the operational semantics. 
The labeled transition system can be deduced from this encoding.}
State of the system consists of instruction pointer
and a map from registers to their values, 
as shown in \cref{fig:regmachine-thrdstate}. 
The rules of the semantics are standard 
(see \cref{fig:regmachine-thrdsem}).

>>>>

For our case study we use a simple idealised model 
of a register machine, which consists of a finite 
sequence of instructions, an instruction pointer 
and an infinite set of registers. 
The syntax of the machine's language is 
shown on~\cref{fig:regmachine-syntax}.

@@

The semantics is given in a form of 
\emph{labelled transition system}~%
\footnote{As we have mentioned, in our \coq development we actually use 
the monadic encoding of the operational semantics. 
The labelled transition system can be deduced from this encoding.}
State of the system consists of instruction pointer
and a map from registers to their values, 
as shown in \cref{fig:regmachine-thrdstate}. 
The rules of the semantics are standard 
(see \cref{fig:regmachine-thrdsem}).

### 2021-04-08 papers/SYRCoSE/related.tex
Typos and British spelling
find . -name "*.tex" -maxdepth 1 -exec aspell --lang=en_GB --mode=tex check "{}" \;

Event structures were introduced by Winskel to study the semantics of 
calculus of communicating systems~\cite{Winskel:82}. 
Several modification of event structures~\cite{Langerak:91, Boudol-Castellani:1991}
were later proposed to tackle similar problems.  
To the best of our knowledge, there were no attempts 
to mechanize the classical theory of event structures 
in proof assistants.

More recently event structures were applied 
in the context of relaxed memory models~%
\cite{Jeffrey-Riely:LICS16, PichonPharabod-Sewell:POPL16, Chakraborty-Vafeiadis:POPL19, Moiseenko-al:ECOOP20}.
Among this line of work, we are aware of only one paper~\cite{Moiseenko-al:ECOOP20}
that was accompanied with the mechanization in a proof assistant. 
The authors formalized \weakestmo~\cite{Chakraborty-Vafeiadis:POPL19} 
memory memory model in \coq. However, this memory model uses 
custom variant of event structures, that does not 
obey axioms of any conventional class of 
event structures~\cite{Winskel:82, Langerak:91, Boudol-Castellani:1991}. 
This fact makes it harder to reuse and adapt it 
for other applications of the theory.

>>>>

Event structures were introduced by Winskel to study the semantics of 
calculus of communicating systems~\cite{Winskel:82}. 
Several modification of event structures~\cite{Langerak:91, Boudol-Castellani:1991}
were later proposed to tackle similar problems.  
To the best of our knowledge, there were no attempts 
to mechanise the classical theory of event structures 
in proof assistants.

More recently event structures were applied 
in the context of relaxed memory models~%
\cite{Jeffrey-Riely:LICS16, PichonPharabod-Sewell:POPL16, Chakraborty-Vafeiadis:POPL19, Moiseenko-al:ECOOP20}.
Among this line of work, we are aware of only one paper~\cite{Moiseenko-al:ECOOP20}
that was accompanied with the mechanisation in a proof assistant. 
The authors formalised \weakestmo~\cite{Chakraborty-Vafeiadis:POPL19} 
memory memory model in \coq. However, this memory model uses 
custom variant of event structures, that does not 
obey axioms of any conventional class of 
event structures~\cite{Winskel:82, Langerak:91, Boudol-Castellani:1991}. 
This fact makes it harder to reuse and adapt it 
for other applications of the theory.
